---
title: "Practical Machine Learning"
author: "Bernd Schnabl"
date: "Tuesday, October 22, 2014"
output: html_document
---


##- Data import and cleaning

The training dataset contains many values that are not present, 
these are either blank "", NA, or DIV/0.
While importing the raw csv files, these values are replaced with NA 
and columns that contain at least one occurance of an NA values are removed in a second step. 

A manual inspection of the data shows, that the first 7 columns contain data, which are not usable for predicting the outcome "classe". These columns are: "X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window".


```{r, echo=TRUE,warning=FALSE,message=FALSE}
setwd("~/R/Coursera/DSSpec/assignments/C08-P01")

read.dataset <- function( fn ) {
  dsraw  <- read.csv(fn,header=T, strip.white=T, na.strings=c("NA","","DIV/0!"), sep=",")
  dsraw[,8:ncol(dsraw)] ## remove the first 7 columns
}

dsTestingFull <- read.dataset("pml-testing.csv")
dsTrainingFull <- read.dataset("pml-training.csv")

validColumnsTr <- ! apply( dsTrainingFull , 2 , function(x) any(is.na(x)) )
validColumnsTst <- ! apply( dsTestingFull , 2 , function(x) any(is.na(x)) )

dsTraining <- dsTrainingFull[ , names( validColumnsTr[validColumnsTr==TRUE])]
dsTesting  <- dsTestingFull[ , names( validColumnsTst[validColumnsTst==TRUE])]
dim(dsTraining)

```


The variable to predict is the classe variable. This variable has 5 levels, where A is the best and E ist the least perfect conducted training.

```{r, echo=TRUE,warning=FALSE,message=FALSE}
unique(dsTrainingFull$classe)
```


A correlation plot can be used to determine, if any features should be excluded, 
since they have a perfect negative or positive correlation.

```{r, echo=TRUE,warning=FALSE,message=FALSE}
library(corrplot)
M <- cor(subset(dsTraining, select=-c(classe)))
corrplot(M, order="hclust", sig.level = 0.0051)
```



## Random Forest Model

The downloaded training data set is divided into 2 partition based on the outcome variable "classe". 


```{r, echo=TRUE,warning=FALSE,message=FALSE}
library(caret)
library(randomForest)

inTrain <- createDataPartition(dsTraining$classe, p=0.7, list=F)
dsTraining_train <- dsTraining[inTrain,]
dsTraining_test <- dsTraining[-inTrain,]

TrainCtrlObj <- trainControl(
               method = "cv", 
               repeats =  1,
               p = 0.75, 
               horizon = 1,
               returnData = TRUE,
               summaryFunction = defaultSummary,
               selectionFunction = "best",
               allowParallel = TRUE)

fit.rf <- train(
      classe ~ ., # predict classe wird all features
      data=dsTraining_train,
      method = "rf", # random forest
      preProcess = c("center", "scale", "knnImpute"), # principle component analysis "pca"
      #metric = ifelse(is.factor(y), "Accuracy", "RMSE"),   
      #maximize = ifelse(metric == "RMSE", FALSE, TRUE),
      trControl = TrainCtrlObj) ## from TrainControl

```


## Print the importance of the variables
```{r, echo=TRUE,warning=FALSE,message=FALSE}
plot(varImp(fit.rf, scale = T),top=20)
```
(Knitr does not correctly display the variable names. The separately created plot does show the correct names on the 20 most relevant features)

## Predicting out of sample errors

The calculation of the root mean square error on the training set delivers:
```{r, echo=TRUE,warning=FALSE,message=FALSE}

predValues_train <- predict(fit.rf, newdata=dsTraining_train)
cm_train <- confusionMatrix(predValues_train, factor(dsTraining_train$classe))
cm_train

```

The accuracy of the ur of sample errors needs to be lower tham the 
accuracy of the predictions on the training set.

```{r, echo=TRUE,warning=FALSE,message=FALSE}

predValues_test <- predict(fit.rf, newdata=dsTraining_test)
cm_test <- confusionMatrix(predValues_test, factor(dsTraining_test$classe))
cm_test

```

## Predicting the Results 

The predictions on the testing dataset with the 20 samples based on the random forest model delivers these results

```{r, echo=TRUE,warning=FALSE,message=FALSE}
res <- predict(fit.rf, dsTesting)
res
```

